---
title: "Worked Example: Simulate from a Gaussian Mixture"
author: "Jack Baker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian Mixture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this example we use the package to infer the modes of a bimodal, 2d Gaussian using [stochastic gradient Hamiltonian Monte Carlo](https://arxiv.org/pdf/1402.4102v2.pdf). So we assume we have independent and identically distributed data $x_1, \dots, x_N$ with $X_i | \theta \sim 0.5 N( \theta_1, I_2 ) + 0.5 N( \theta_2, I_2 )$, and we want to infer $\theta_1$ and $\theta_2$.

First, let's simulate the data with the following code, we set $N$ to be $10^4$
```{r message=FALSE}
library(sgmcmc)
library(MASS)
# Declare number of observations
N = 10^4
# Set locations of two modes, theta1 and theta2
theta1 = c( 0, 0 )
theta2 = c( 0.1, 0.1 )
# Allocate observations to each component
set.seed(13)
z = sample( 2, N, replace = TRUE, prob = c( 0.5, 0.5 ) )
# Predeclare data matrix
X = matrix( rep( NA, 2*N ), ncol = 2 )
# Simulate each observation depending on the component its been allocated
for ( i in 1:N ) {
    if ( z[i] == 1 ) {
        X[i,] = mvrnorm( 1, theta1, diag(2) )
    } else {
        X[i,] = mvrnorm( 1, theta2, diag(2) )
    }
}
dataset = list("X" = X)
```
In the last line we defined the dataset as it will be input to the relevant `sgmcmc` function. A lot of the inputs to functions in `sgmcmc` are defined as lists. This improves flexibility by enabling models to be specified with multiple parameters, datasets and allows separate tuning constants to be set for each parameter. We assume that observations are always accessed on the first dimension of each object, i.e. the point $x_i$ is located at `X[i,]` rather than `X[,i]`. Similarly the observation $i$ from a 3d object `Y` would be located at `Y[i,,]`.

The parameters are declared very similarly, but this time the value associated with each entry is its starting point. We have two parameters `theta1` and `theta2`, which we'll just start from the true values for the sake of demonstration purposes
```{r}
params = list( "theta1" = c( 0, 0 ), "theta2" = c( 0.1, 0.1 ) )
```

Now we'll define the functions `logLik` and `logPrior`. It should now become clear why the list names come in handy. The function `logLik` should take two parameters as input: `params` and `dataset`. These parameters will be lists with the same names as those you defined for `params` and `dataset` earlier. There is one difference though, the objects in the lists will have automatically been converted to `TensorFlow` objects for you. The `params` list will contain `TensorFlow` tensor variables; the `dataset` list will contain `TensorFlow` placeholders. The `logLik` function should take these lists as input and return the value of the log likelihood as a tensor at point `params` given data `dataset`. The function should do this using `TensorFlow` operations, as this allows the gradient to be automatically calculated; it also allows the wide range of distribution objects as well as matrix operations that `TensorFlow` provides to be taken advantage of. A tutorial of `TensorFlow` for `R` is beyond the scope of this article, for more details we refer the reader to the website of [TensorFlow for R](https://tensorflow.rstudio.com/). With this in place we can define the log-likelihood function `logLik` as follows
```{r}
logLik = function( params, dataset ) {
    # Declare Sigma (assumed known)
    SigmaDiag = c(1, 1)
    # Declare distribution of each component
    component1 = tf$contrib$distributions$MultivariateNormalDiag( params$theta1, SigmaDiag )
    component2 = tf$contrib$distributions$MultivariateNormalDiag( params$theta2, SigmaDiag )
    # Declare allocation probabilities of each component
    probs = tf$contrib$distributions$Categorical(c(0.5,0.5))
    # Declare full mixture distribution given components and allocation probabilities
    distn = tf$contrib$distributions$Mixture(probs, list(component1, component2))
    # Declare log likelihood
    logLik = tf$reduce_sum( distn$log_prob(dataset$X) )
    return( logLik )
}
```
So this function basically states that our log-likelihood function is $\sum_{i=1}^N \log \left[ 0.5 \mathcal N( x_i | \theta_1, I_2 ) + 0.5 \mathcal N( x_i | \theta_2, I_2 ) \right]$, where $\mathcal N( x | \mu, \Sigma )$ is a Gaussian density at $x$ with mean $\mu$ and variance $\Sigma$. Most of the time just specifying the constants in these functions, such as `SigmaDiag`, as `R` objects will be fine. But there are sometimes issues when these constants get automatically converted to `tf$float64` objects by `TensorFlow` rather than `tf$float32`. If you run into errors involving `tf$float64` then force the constants to be input as `tf$float32` by using `SigmaDiag = tf$constant( c( 1, 1 ), dtype = tf$float32 )`.

Next we want to define our log-prior density, which we assume is uninformative $\log p( \theta ) = \log \mathcal N(\theta | 0,10I_2)$. Similar to the log-likelihood function, the log-prior density is defined as a function, but only with input `params`. In our case the definition is
```{r}
logPrior = function( params ) {
    # Declare hyperparameters mu0 and Sigma0
    mu0 = c( 0, 0 )
    Sigma0Diag = c(10, 10)
    # Declare prior distribution
    priorDistn = tf$contrib$distributions$MultivariateNormalDiag( mu0, Sigma0Diag )
    # Declare log prior density and return
    logPrior = priorDistn$log_prob( params$theta1 ) + priorDistn$log_prob( params$theta2 )
    return( logPrior )
}
```

Finally we set the tuning parameters for SGHMC, this is a list with the same names as the `params` list you defined earlier, and values are the stepsize for that parameter.
```{r}
stepsize = list( "theta1" = 2e-5, "theta2" = 2e-5 )
```
Optionally, we can set the tuning parameter for the momentum `alpha` in the same way as the `stepsize`. But we'll leave this along with the trajectory tuning constant `L`, and the `minibatchSize` as their defaults.

Now we can run our SGHMC algorithm using the `sgmcmc` function `sghmc`, which returns a list of Markov chains for each parameter as output. Use the argument `verbose = FALSE` to hide the output of the function. To make the results reproducible we'll set the seed to 13. We'll set the number of iterations as `11000` to allow for `1000` iterations of burn-in.
```{r eval=FALSE}
chains = sghmc( logLik, dataset, params, stepsize, logPrior = logPrior, nIters = 11000, 
        verbose = FALSE, seed = 13 )
```

```{r echo=FALSE}
tryCatch({
    chains = sghmc( logLik, dataset, params, stepsize, logPrior = logPrior, nIters = 11000, 
            verbose = FALSE, seed = 13 )
}, error = function (e) { 
    writeLines("Not all tensorflow dependencies are met so skipping this...")
    writeLines("Try running tensorflow::install_tensorflow().")
})
```

Finally, we'll plot the results after removing burn-in
```{r eval=FALSE}
library(ggplot2)
# Remove burn in
burnIn = 10^3
chains = list( "theta1" = as.data.frame( chains$theta1[-c(1:burnIn),] ),
               "theta2" = as.data.frame( chains$theta2[-c(1:burnIn),] ) )
# Concatenate the two chains for the plot to get a picture of the whole distribution
plotData = rbind(chains$theta1, chains$theta2)
ggplot( plotData, aes( x = V1, y = V2 ) ) +
    stat_density2d( size = 1.5, alpha = 0.7 )
```

```{r echo=FALSE}
tryCatch({
    a = tf$constant(c(1, 1))
    library(ggplot2)
    # Remove burn in
    burnIn = 10^3
    chains = list( "theta1" = as.data.frame( chains$theta1[-c(1:burnIn),] ),
            "theta2" = as.data.frame( chains$theta2[-c(1:burnIn),] ) )
    # Concatenate the two chains for the plot to get a picture of the whole distribution
    plotData = rbind(chains$theta1, chains$theta2)
    ggplot( plotData, aes( x = V1, y = V2 ) ) +
            stat_density2d( size = 1.5, alpha = 0.7 )
}, error = function (e) { 
    writeLines("Not all tensorflow dependencies are met so skipping this...")
    writeLines("Try running tensorflow::install_tensorflow().")
})
```
